Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-06-14 cmsb22.tex
Observations

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the official PyBoolNet repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}, the results are comparable with all minimal trap spaces found very fast.
\revise{For 5 of the 29 models, mpbn does not give any answer since these models are not locally-monotonic.
Note that on some very small models, Trappist is sometimes slower than PyBoolNet and/or mpbn, but still significantly under one second.}
Moreover, we believe that the result on the arellano\_rootstem model is caused by the cold start of the JVM for BioLQM.\@
\revise{On the contrary, on every model that was a bit challenging for PyBoolNet or mpbn, the new method is far more efficient with speedups between one and two orders of magnitude.}

@@

%%% The proposed method outperforms PyBoolNet %%%
The first observation is that the proposed method vastly outperforms PyBoolNet in computational time, on each and every model, and sometimes with orders of magnitude of difference (e.g., for most models in the 100--1000 nodes size range).
Note that for all the cases where PyBoolNet did not manage to finish before the timeout, as marked by ``DNF'' in Table~\ref{tab:result_real}, the timeout occurred during the computation of the prime-implicants.
Hence, not even a single minimal trap space was output by that method.
The computational advantage is therefore immediately a practical advantage since on the one hand the state-of-the-art method did not allow any analysis whatsoever of the models, and on the other hand the proposed method could provide, very often under one second, the first thousand minimal trap spaces.
For modellers having a critical look at a model and in a \emph{model, invalidate, refine} loop this means a huge difference in the models that are amenable to study.

@@

%%% It is not clear to show the advantage in ASP encoding %%%
The second observation is that for all the models where PyBoolNet finished before the timeout, once PyBoolNet went through the prime-implicant phase, its ASP solving phase quickly returned the first 1000 minimal trap spaces, all under one second.
For these models, the ASP solving phase of the proposed method also took very short time, all under one second.
Hence, with the experimental results shown in this paper, the practical differences between our ASP encoding and that of PyBoolNet are not distinctly exposed.
The fact that our new ASP encoding is guaranteed to be linear in the number of nodes of the original model does not seem to be crucial here, however a much deeper analysis of those cases remains to be done.

%%% For very huge models %%%
The third observation is that for very large models (i.e., more than two thousand nodes) the proposed method did not manage to finish the Petri net conversion before the timeout, as marked by ``???'' in Table~\ref{tab:result_real}.
This points to the fact that our current choice of using a BDD-based translation to obtain that Petri net encoding, though it provides a small/efficient ASP might be too costly to handle the largest models.
In such a case, a more \emph{naive} encoding might provide a much larger ASP program, with many redundant rules, but easier/faster to obtain.
The evaluation of the feasibility of such strategy, and of its impact on smaller instances, remains to be done and is out of the scope of this first article.

>>>>

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the official PyBoolNet repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}, the results are comparable with all minimal trap spaces found very fast.
\revise{For 5 of the 29 models, mpbn did not give any answer because it recognized that these models are not locally-monotonic.
Note that on some very small models, Trappist is sometimes slower than PyBoolNet and/or mpbn, but still significantly under one second.}
Moreover, we believe that the result on the arellano\_rootstem model is caused by the cold start of the JVM for BioLQM.\@
\revise{On the contrary, on every model that was a bit challenging for PyBoolNet or mpbn, the new method is far more efficient with speedups between one and two orders of magnitude.}

@@

%%% The proposed method and mpbn %%%
\revise{The first observation is that for 27 of the 33 models (more than 78\%), mpbn did not give any answer because it recognized that these models are not locally-monotonic.
For 6 of the 33 models where mpbn returned the answers, mpbn and Trappist are comparable in computation time.
Along with the fact that mbpn can handle only locally-monotonic models and Trappist can handle general models, it is less insightful to further compare between them.
Hence, we focus on only comparisons between PyBoolNet and Trappist in the following observations.}

%%% The proposed method outperforms PyBoolNet %%%
\revise{The second observation} is that the proposed method vastly outperforms PyBoolNet in computational time, on each and every model, and sometimes with orders of magnitude of difference (e.g., for most models in the 100--1000 nodes size range).
Note that for all the cases where PyBoolNet did not manage to finish before the timeout, as marked by ``DNF'' in Table~\ref{tab:result_real}, the timeout occurred during the computation of the prime-implicants.
Hence, not even a single minimal trap space was output by that method.
The computational advantage is therefore immediately a practical advantage since on the one hand the state-of-the-art method did not allow any analysis whatsoever of the models, and on the other hand the proposed method could provide, very often under one second, the first thousand minimal trap spaces.
For modellers having a critical look at a model and in a \emph{model, invalidate, refine} loop this means a huge difference in the models that are amenable to study.

@@

%%% It is not clear to show the advantage in ASP encoding %%%
\revise{The third observation} is that for all the models where PyBoolNet finished before the timeout, once PyBoolNet went through the prime-implicant phase, its ASP solving phase quickly returned the first 1000 minimal trap spaces, all under one second.
For these models, the ASP solving phase of the proposed method also took very short time, all under one second.
Hence, with the experimental results shown in this paper, the practical differences between our ASP encoding and that of PyBoolNet are not distinctly exposed.
The fact that our new ASP encoding is guaranteed to be linear in the number of nodes of the original model does not seem to be crucial here, however a much deeper analysis of those cases remains to be done.

%%% For very huge models %%%
\revise{The last observation} is that for very large models (i.e., more than two thousand nodes) the proposed method did not manage to finish the Petri net conversion before the timeout, as marked by ``???'' in Table~\ref{tab:result_real}.
This points to the fact that our current choice of using a BDD-based translation to obtain that Petri net encoding, though it provides a small/efficient ASP might be too costly to handle the largest models.
In such a case, a more \emph{naive} encoding might provide a much larger ASP program, with many redundant rules, but easier/faster to obtain.
The evaluation of the feasibility of such strategy, and of its impact on smaller instances, remains to be done and is out of the scope of this first article.
